---
title: "Lab4 Individual/Team"
author: "Colin Willenborg"
date: "10/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyr)
library(dplyr)
library(ggplot2)
defense <- read_excel('./cyclonesFootball2019.xlsx', sheet='Defensive')
offense <- read_excel('./cyclonesFootball2019.xlsx', sheet='Offensive')
biography <- read_excel('./cyclonesFootball2019.xlsx', sheet='Biography')
```

## Cleaning the data

```{r}
offClean <- offense %>%
  mutate_at(vars(Name,
            Opponent_Opponent),
            list(factor)) %>%
  separate('Passing_CMP-ATT', c('Passing_CMP', 'Passing_ATT'))%>%
  mutate(across(.cols = (Rushing_ATT:Passing_INT), .fns = (as.numeric)))

defClean <- defense %>%
  mutate_at(vars(Name,
            Opponent_Opponent),
            list(factor)) %>%
  mutate(across(.cols = (Tackles_Solo:Pass_PB), .fns = (as.numeric)))

bioClean <- biography %>%
  mutate_at(vars(Name),
            list(factor)) %>%
  mutate(Weight = as.numeric(Weight)) %>%
  separate(Height, c('feet', 'inches'), sep='-') %>%
  mutate(feet = as.numeric(feet), inches = as.numeric(inches)) %>%
  mutate(Height_inches = (12 * feet + inches)) %>%
  select(-feet, -inches)

str(offClean)
```

```{r}
str(defClean)
```

```{r}
str(bioClean)
```

## Part 2: Tidying the data

```{r}
defCleanPivot <- pivot_longer(defClean, Tackles_Solo:Pass_PB, names_to ="Defense_Skill", values_to = "stat")
skills <- c('Tackles_Solo',	'Tackles_ASST',	'Tackles_TFL',	'Tackles_Sack',	'Turnover_FF', 'Turnover_FR',	'Turnover_INT',	'Pass_QBH	Pass_PB')
ggplot(defCleanPivot, aes(x=Defense_Skill, y=stat)) + 
  geom_bar(stat= 'identity') + scale_x_discrete(limits = skills) + 
  theme(axis.text.x = element_text(angle = 90)) + ylab("Count") + 
  xlab("Defensive Skill")

```

The three rarest defensive skills in this dataset are fumble, fumble recovery, and interception.

```{r}
defvsIowa <- defCleanPivot %>%
  filter(Opponent_Opponent == c('Iowa', 'Notre Dame'),
         Defense_Skill == 'Tackles_Solo')
ggplot(defvsIowa, aes(x=Name, y=stat, color=Opponent_Opponent)) + 
  geom_point() + theme(axis.text.x = element_text(angle = 90)) + labs(color='Opponent')
```

Iowa State defense had a total of 21 tackles against Notre Dame and a total of 8 against Iowa so we had a better defense against Notre Dame. We also had more players with larger amounts of tackles with 3 players above 4 tackles against Notre Dame and nobody above 2 tackles against Iowa.

```{r}
bioClean <- bioClean %>%
  separate(Hometown, c('City', 'State'), sep=',')
bioCleanState <- bioClean %>% 
  select(State)
tableBio <- table(bioCleanState)
tableBio
```
## Part 3: Joining Data Frames

```{r}
offenseState <- left_join(offClean, bioClean, by='Name')
offenseState <- offenseState %>% 
  select(Name, State) %>% 
  unique()
offenseState <-offenseState %>% 
  select(State)
offenseState <- table(offenseState)
offenseState
```

```{r}
defenseState <- inner_join(defClean, bioClean, by='Name')
defenseState <- defenseState %>% 
  select(Name, State) %>% 
  unique()
defenseState
defenseState <- defenseState %>% 
  select(State)
defenseStateTable <- table(defenseState)
defenseStateTable
```
For the offense data set there are 4 players who have no Biography data so wanted to say that here before going into the analysis. The offense has fewer players to look at but it is very spread out with the most of 2 being from one state both from Florida and Oklahoma. On the other hand defense has multiple states with a lot of players from them and has 7 home grown players from Iowa. 

```{r}
wins <- c('UNI', 'ULM', 'TCU', 'West Virginia', 'Texas Tech', 'Texas', 'Kansas')
offCleanPurdy <- offClean %>%
  filter(Name == "Purdy, Brock") %>%
  select(-Receiving_REC, -Receiving_TD, -Receiving_YDS) %>% 
  mutate(QB_Rating = (Passing_CMP/Passing_ATT) + (3.3 * (Passing_TD/Passing_ATT) -(2*(Passing_INT/Passing_ATT)) + (8.4 * Passing_YDS/Passing_ATT))) %>% 
  mutate(Win = ifelse(Opponent_Opponent %in% wins, 'W', 'L')) %>% 
  arrange(QB_Rating) %>%
  select(Name, Opponent_Opponent, QB_Rating, Win)
offCleanPurdy
```

I followed a formula to find the NCAA Quarterback Rating and then arranged the table by that. This shows that when Purdy is performing ISU wins games. He won 5 of the top 7 games for his QB_Rating. So when he performs ISU does well.


Defense analysis

```{r}
defense18 <- read_excel('./cyclonesFootball2018.xlsx', sheet='Defensive')

defClean2018Name <- defense18 %>% select(Name) %>% distinct()
defClean2019Name <- defClean %>% select(Name) %>% distinct()
namesDefense <- inner_join(defClean2019Name, defClean2018Name, by='Name')

defClean2018 <- defense18 %>%
  mutate_at(vars(Name,
            Opponent_Opponent),
            list(factor)) %>%
  mutate(across(.cols = (Tackles_Solo:Pass_PB), .fns = (as.numeric))) %>% 
  group_by(Name) %>% 
  summarise_at(vars(Tackles_Solo:Pass_PB), sum, na.rm=TRUE) %>% 
  mutate(Year = 2018) %>% 
  right_join(namesDefense) %>% 
  pivot_longer('Tackles_Solo':'Pass_PB', names_to='Defense_Skill', values_to='stat')

defClean2019 <- defClean %>% 
  select(-Opponent_Opponent) %>% 
  group_by(Name) %>% 
  summarise_at(vars(Tackles_Solo:Pass_PB), sum, na.rm=TRUE) %>% 
  mutate(Year = 2019) %>% 
  right_join(namesDefense) %>% 
  pivot_longer('Tackles_Solo':'Pass_PB', names_to='Defense_Skill', values_to='stat')

def2018_2019 <- full_join(defClean2018, defClean2019)  

ggplot(def2018_2019, aes(x=Defense_Skill, y=stat, fill=Year)) +
  geom_bar(stat='identity') +
  theme(axis.text.x = element_text(angle=90)) +
  facet_wrap(~Name, 6)
```

```{r}
def2018_2019new <- def2018_2019[def2018_2019$Name == 'Bailey, JaQuan' | def2018_2019$Name == 'Eisworth, Greg' | def2018_2019$Name == 'Lewis, Braxton' | 
                             def2018_2019$Name == 'Hummel, Jake' | def2018_2019$Name == 'Johnson, Anthony' | def2018_2019$Name == 'Rose, Mike' |
                             def2018_2019$Name == 'Uwazurike, Eyioma' | def2018_2019$Name == "Vance, O'Rien" | def2018_2019$Name == 'White, Lawrence',]
def2018_2019new
ggplot(def2018_2019new, aes(x=Defense_Skill, y=stat, fill=Year)) +
  geom_bar(stat='identity') +
  theme(axis.text.x = element_text(angle=90)) +
  facet_wrap(~Name)
```
For the defense players I first looked at all the players and then I took the biggest graphs from that and put them in a new data frame to analize them easier. This shows that the most improved player defensively would have been between Mike Rose, Lawerence White, and Greg Eisworth.

```{r}
offense18 <- read_excel('./cyclonesFootball2018.xlsx', sheet='Offensive')

offClean2018Name <- offense18 %>% select(Name) %>% distinct()
offClean2019Name <- offClean %>% select(Name) %>% distinct()
namesOffense <- inner_join(offClean2018Name, offClean2019Name, by='Name')

offClean2018 <- offense18 %>%
  mutate_at(vars(Name,
            Opponent_Opponent),
            list(factor)) %>%
  mutate(across(.cols = (Rushing_ATT:Receiving_TD), .fns = (as.numeric))) %>% 
  group_by(Name) %>% 
  summarise_at(vars(Rushing_ATT:Receiving_TD), sum, na.rm=TRUE) %>% 
  mutate(Year = 2018) %>% 
  filter(Name != 'Purdy, Brock') %>% 
  right_join(namesOffense) %>% 
  pivot_longer('Rushing_ATT':'Receiving_TD', names_to='Offense_Skill', values_to='stat')

offClean2019 <- offClean %>% 
  select(-Opponent_Opponent) %>% 
  group_by(Name) %>% 
  summarise_at(vars(Rushing_ATT:Receiving_TD), sum, na.rm=TRUE) %>% 
  mutate(Year = 2019) %>% 
  filter(Name != 'Purdy, Brock') %>% 
  right_join(namesOffense) %>% 
  pivot_longer('Rushing_ATT':'Receiving_TD', names_to='Offense_Skill', values_to='stat')

off2018_2019 <- full_join(offClean2018, offClean2019)  

ggplot(off2018_2019, aes(x=Offense_Skill, y=stat, fill=Year)) +
  geom_bar(stat='identity') +
  theme(axis.text.x = element_text(angle=90)) +
  facet_wrap(~Name)
```

```{r}
offClean2018 <- offense18 %>%
  mutate_at(vars(Name,
            Opponent_Opponent),
            list(factor)) %>%
  separate('Passing_CMP-ATT', c('Passing_CMP', 'Passing_ATT'))%>%
  mutate(across(.cols = ('Passing_CMP':Passing_INT), .fns = (as.numeric))) %>% 
  group_by(Name) %>% 
  summarise_at(vars('Passing_CMP':Passing_INT), sum, na.rm=TRUE) %>% 
  mutate(Year = 2018) %>% 
  right_join(namesOffense) %>% 
  pivot_longer('Passing_CMP':'Passing_INT', names_to='Offense_Skill', values_to='stat')

offClean2019 <- offClean %>% 
  select(-Opponent_Opponent) %>% 
  group_by(Name) %>% 
  summarise_at(vars('Passing_CMP':Passing_INT), sum, na.rm=TRUE) %>% 
  mutate(Year = 2019) %>% 
  right_join(namesOffense) %>% 
  pivot_longer('Passing_CMP':'Passing_INT', names_to='Offense_Skill', values_to='stat')

off2018_2019 <- full_join(offClean2018, offClean2019)  

off2018_2019 <- off2018_2019[Name="Purdy, Brock", ]

ggplot(off2018_2019, aes(x=Offense_Skill, y=stat, fill=Year)) +
  geom_bar(stat='identity') +
  theme(axis.text.x = element_text(angle=90)) 
```
For the offense I looked at both the offensive players not including Purdy and then looked at Purdy seperately. It shows that Purdy improved a lot from 2018 to 2019 and is one of the most improved players but some of the other players also improved a lot. Deshaunte Jones is an example of a really big improvement from 2018 barely playing to being one of our best receivers in the 2019 season. 

